---
title: "Movie Lens"
Name: "Hannah Gable"
date: "05/20/2024"
output: word_document:
  toc: yes
  html_document:
    toc: yes
    pdf_document:
      toc:yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#Introduction
  The data set known as the Movie Lens data set was collected on the behalf of Group Lens Research. Are we able to predict movie ratings based upon user preference and/or the age of a movie? Using this data set and the property of least squares, the R script is able to calculate the RMSE based upon Movie ID, age of the movie, and the user ratings.
  
  The Movie Lens data set contains 10677 movies, 69878 users, 797 genres, and 10000054 rows.
  
  The steps I performed for the analysis of the data:
    -Created an age for the movie column
    -Made a graphic display of movie, users and ratings in order to find a pattern of the data
    -Looked into the determination coefficient of R-squared
    -Graphically examined the linear correlation and r-value
    -Calculate the RMSE based upon Movie ID, the age of the movie, and user ID
  After evaluating the movies through their graphical representations and calculating the RMSE, I concluded that the best predictor for the movie ratings was the Movie ID, User ID. The age of the movie had no change or effect on RMSE.
  
  The overall RMSE is 0.8252
  
  ```{r,include=FALSE}
  
  library(stringr)
  library(ggplot2)
  library(caret)
  library(readr)
  library(dplyr)
  library(stats)
  library(corrplot)
  library(tidyr)
  library(lubridate)
  library(data.table)
  library(DT)
  library(knitr)
  library(methods)
  library(gridExtra)
  library(grid)
  library(corrplot)
  library(Matrix)
  library(corrplot)
  library(RColorBrewer)
  library(magrittr)
  library(stringi)
  library(viridis)
  library(heuristica)
  library(MatrixStats)
  library(gam)
  library(modelr)
  library(tidyselect)
  library(tidyr)
  library(tibble)
  library(purr)
  library(broom)
  library(forcats)
  library(DBI)
  library(hexbin)
  
  ```
#How to download the data
  ```{r,echo=TRUE}
  dl <- tempfile()
  download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip",dl)
  ratings <- read.table(text=gsub("::","\t",readLines(unzip(dl,-10m100K/ratings.dat"))),
  col.names=c("userID","movieID","rating","timestamp"))
  ```
  
#Building the data set
  ```{r}
  movies<-str_split_fixed(readLines(unzip(dl,"ml-10M100K/movies.dat")),"\\::",3)
  colnames(movies)<-c("movieid","title","genres")
  movies<-as.data.frame(movies) %>% mutate(movieid=as.numeric(levels(movieid))[movieid],title=as.character(title),genres=as.character(genres))
  
  #Evaluate the size of the data
movielens<-left_join(ratings,movies,by="movieid")
nrow(movielens)
n_distinct(movielens$movieid)
n_distinct(movielens$genres)
n_distinct(movielens$userid)
```

#Validation of 10% of the Movie Lens data 
```{r}
set.seed(1)
test_index<-createDataPartition(y=movielens$rating,times=1,p=0.1,list=FALSE)
edx<-movielens[-test_index,]
temp<-movielens[test_index,]
```

#userid and movieid in the edx data set
```{r}
validation<-temp%>% semi_join(edx,by="movieid")%>% semi_join(edx,by="userid")
```
#Put rows back into edx data set that were once removed
```{r}
removed<-anti_join(temp,validation)
edx<-rbind(edx,removed)
```
#data cleaning, evaluation, and visualization
#Is the age of the movie a factor for predicting the rating of the movie? I took the movie premier date and calculated the age of the movie. I will look into genres and the effects of the ratings.
```{r}
head(edx)
glimpse(edx)
```
#Number of distinct users, genres, and movies
```{r}
n_distinct(edx$userid)
n_distinct(edx$genres)
n_distinct(edx$movieid)
nrow(edx)
```

#Converting the Timestamp to the year
```{r}
edx<-mutate(edx,year_rated=year(as_datetime(timestamp)))
head(edx)
```

#Extract the date of the movie premier and calculate the age of the movie. Does the age of the movie predict the ratings?
```{r}
premier<-stringi::stri_extract(edx$title,regex="(\\d{4})",comments=TRUE)%>%as.numeric()
edx_with_title_dates<-edx%>%mutate(premier_date=premier)
head(edx_with_title_dates)
```
#Accuracy check
```{r}
edx_with_title_dates<-edx_with_title_dates%>%select(-timestamp)
head(edx_with_title_dates)

#Are the dates correct?
edx_with_title_dates%>%filter(premier_date>2018)%>%group_by(movieid,title,premier_date)%>%summarize(n=n())
edx_with_title_dates%>%filter(premier_date<1900)%>%group_by(movieid,title,premier_date)%>%summarize(n=n())

#Fix the dates that are displayed incorrectly
edx_with_title_dates[edx_with_title_dates$movieid=="27266","premier_date"]<-2004
edx_with_title_dates[edx_with_title_dates$movieid=="671","premier_date"]<-1996
edx_with_title_dates[edx_with_title_dates$movieid=="2308","premier_date"]<-1973
edx_with_title_dates[edx_with_title_dates$movieid=="4159","premier_date"]<-2001
edx_with_title_dates[edx_with_title_dates$movieid=="5310","premier_date"]<-1985
edx_with_title_dates[edx_with_title_dates$movieid=="8864","premier_date"]<-2004
edx_with_title_dates[edx_with_title_dates$movieid=="1422","premier_date"]<-1997
edx_with_title_dates[edx_with_title_dates$movieid=="4311","premier_date"]<-1998
edx_with_title_dates[edx_with_title_dates$movieid=="5472","premier_date"]<-1972
edx_with_title_dates[edx_with_title_dates$movieid=="6290","premier_date"]<-2003
edx_with_title_dates[edx_with_title_dates$movieid=="6645","premier_date"]<-1971
edx_with_title_dates[edx_with_title_dates$movieid=="8198","premier_date"]<-1960
edx_with_title_dates[edx_with_title_dates$movieid=="8905","premier_date"]<-1992
edx_with_title_dates[edx_with_title_dates$movieid=="53953","premier_date"]<-2007
```

#Age of the movie
```{r}
edx_with_title_dates<-edx_with_title_dates%>%mutate(age_of_movie=2018-premier_date,rating_date_range=year_rated-premier_date)
head(edx_with_title_dates)
```
#Graphing the age of the movie
```{r,message=FALSE}
edx%>%group_by(movieid)%>%summarize(n=n()%>%ggplot(aes(n))+geom_histogram(fill="cadetblue3",color="grey20",bins=10)+scale_x_log10()+ggtitle("Number of Movies Ratings")
```
#Finding the number of ratings by the userid
```{r,message=FALSE}
edx%>%group_by(userid)%>%summarize(n=n())%>%ggplot(aes(n))+geom_histogram(fill="cadetblue3",color="grey20",bins=10)+scale_x_log10()+ggtitle("Number of Users Ratings")
```
#Calculate the average of the user rating, movie rating, age of the movie, and the year
```{r}
user_avgs<-edx_with_title_dates%>%group_by(userid)%>%summarize(avg_user_rating=mean(rating))
movie_avgs<-edx_with_title_dates%>%group_by(movieid)%>%summarize(avg_movie_rating=mean(rating))
age_avgs<-edx_with_title_dates%>%group_by(age_of_movie)%>%summarize(avg_rating_by_age=mean(rating))
year_avgs<-edx_with_title_dates%>%group_by(year_rated)%>%summarize(avg_rating_by_year=mean(rating))
head(age_avgs)
```
```{r}
head(user_avgs)
```
#What is the graphical relationship between the average of the movie ratings and the age of the movie?
#This graph demonstrates older movies having higher ratings, then those ratings dropping.
```{r,message=FALSE}
age_avgs%>%ggplot(aes(age_of_movie,avg_rating_by_age))+geom_point()+ggtitle("Age of the movie vs. Average of the movie ratings")
```
#This graph will demonstrate to us that the average ratings are consistent between the values of 2.5 and 4.5
```{r,message=FALSE}
user_avgs%>%ggplot(aes(userid,avg_user_rating))+geom_point(alpha=1/20,colour="blue")+ggtitle("Users vs. The Average User Rating")
```
#Calculating the lm of the Age of the movie vs. Average of the movie ratings
```{r}
summary(lm(avg_rating_by_age~age_of_the_movie,data=age_avgs))
```
#R-squared is at 0.30

#Plotting the residual values
```{r,message=FALSE}
avg_rating.lm<-lm(avg_rating_by_age~age_of_movie,data=age_avgs)
avg_rating.res<-resid(avg_rating.lm)
plot(age_avgs$age_of_movie,avg_rating.res,ylab='residuals',xlab='age of the movie',main='The average rating by the age of the movie')+abline(0,0)
```
#30% of the movie ratings variation can be predicted by examining if there is a correlation between the rating and the age of the movie
```{r,message=FALSE}
#movies that are younger than 75 years old
age_of_movie_less_than75<-age_avgs%>%filter(age_of_movie<75)
#movies that are younger than 75 years old versus the movie rating average
age_of_movie_less_than75<-ggplot(aes(age_of_movie,avg_rating_by_age))+geom_point()+ggtitle("Age of the movie Vs. Movie Rating Average")
```
#The value of R-squared
```{r}
age_lessthan75_rating.lm<-lm(avg_rating_by_age~age_of_movie,data=age_of_movie_less_than75)
summary(age_lessthan75_rating.lm)
```
#Demonstrates a value of R-squared increasing to 0.745
#Plotting the residual values
```{r,message=FALSE}
head(age_of_movie_less_than75)
age_lessthan75.res<-resid(age_lessthan75_rating.lm)
plot(age_of_movie_less_than75$age_of_movie,age_lessthan75.res,ylab='residuals',xlab='The Age of the Movie',main='Movie Rating Average by the age of the movie')+abline(0,0)
```
#The graph looks more linear when we observe movies in the age range from 20 to 75 years old
```{r}
age_between20_and_75<-age_avgs%>%filter(age_of_movie>20)&(age_of_movie<75))
```
```{r,message=FALSE}
age_between20_and_75%>%ggplot(aes(age_of_movie,avg_rating_by_age))+geompoint()+ggtitle("Movies between the ages of 30 to 75 vs. Movie rating average")
```
#As we can observe, the plot is linear. However, we also observe that the R-squared has decreased to 0.69
```{r}
summary(lm(avg_rating_by_age~age_of_movie,data=age_between20_and_75))
```
#Age of the movies that are between 20 and 40 years of age
```{r}
ages_between20_and_40<-age_avgs%>%filter((age_of_movie<20)&(age_of_movie<40))
age_between20_and_40%>%ggplot(aes(age_of_movie,avg_rating_by_age))+ggeom_point()+ggtitle("The Age of the Movies between 20 and 40 years old")
```
#Older movies are showing to have higher ratings. We can display this graphically
```{r}
summary(lm(avg_rating_by_age~age_of_movie,data=age_between20_and_40))
```
#We can see the R-squared increased to 0.71. We will now explore movies between the ages 0 to 30 years old.
```{r,message=FALSE}
age_less_than30<-age_avgs%>%filter((age_of_movie>30))
#Movies that are less than 30 years old with an average movie rating
age_less_than30%>%ggplot(aes(age_of_movie,avg_rating_by_age))+geom_point()+ggtitle('Age of the Movie less than 30 years old')
```
#We can see that R-squared is closer to 0 with the movies that are less than 30 years old. We can also observe that there is quite a bit of variaration with this one as well
```{r}
summary(lm(avg_rating_by_age~age_of_movie,data=age_less_than30))
```
#The age of the movie does have an effect on the outcome of the average rating of the movie since older movies tend to have those higher ratings
#Do the genres of movies have an effect on the ratings? Genres were split into single genres
```{r}
dat<-edx_with_title_dates%>%separate_rows(genres,sep="\\|")
head(dat)
```
#Number of movies with each genre using movieid
```{r}
genre_count_by_movieid<-dat%>%group_by(movieid,genres)%>%summarize(n=n())
head(genre_count_by_movieid)
```
#Number of Movies in each genre
```{r}
number_of_genres<-dat%>%group_by(genres)%>%summarize(n=n())
number_of_genres
```
#Listing the movie genres
```{r}
genre_list<-number_of_genres$genres
genre_list
```
#Examining the movies ratings distribution
```{r}
temp<-dat%>%group_by(genres)%>%summarize(n=n())%>%ungroup()%>%mutate(sumN=sum(n),percentage=n/sumN)%>%arrange(-percentage)
```
#Creating a bar graph of the movie genres. Here, we can see that the movie genre known as "drama" had the highest rating percentage.
```{r,message=FALSE}
temp%>%ggplot(aes(reorder(genres,percentage),percentage,fill=percentage))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="Y10rRd")+labs(y="percentage",x="genre")+ggtitle("Distributionof Genres by the percentage of movies rates")
```
#Movie genres average rating (the mean). Here, we can see that the movie genre known as "horror" has the lowest movie rating averages and the movie genre known as "film noir" has the highest.
```{r,message=FALSE}
temp<-dat%>%group_by(genres)%>%summarize(mean_rating_by_genre=mean(rating))%>%arrange(-mean_rating_by_genre)
temp%>%ggplot(aes(reorder(mean_rating_by_genre),mean_rating_by_genre,fill=mean_rating_by_genre))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="y10rRd")+labs(y="Mean rating",x="genre")+ggtitle("Average Rating of genres")
```
#Look and observe the ratings of the movies based on the valuation and numerical value of the movie ratings. For this, I took movies with an average that was greater than 4
```{r}
avg_rating_greater_than_4<-edx%>%group_by(title)%>%summarize(mean_rating=mean(rating),n=n())%>%filter(mean_rating>=4)%>%arrange(desc(n=mean_rating))
avg_rating_greater_than_4%>%filter(n>=10000)%>%ggplot(aes(reorder(title,n),n,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+xlab("movie")+ylab("number of ratings")+ggtitle("Movies with an average rating\ngreater than or equal to 4\nand number of ratings>10000")
```
#Observe the ratings of the movies that contain more than 10000 ratings and are between the numbers of 3 and 4
```{r,message=FALSE}
avg_between3_4<-edx%>%group_by(title)%>%summarize(mean_rating=mean(rating),n=n())%>%filter(n>10000,(mean_rating>=3&mean_rating<4))%>%arrange(desc(n,mean_rating))
p<-avg_between3_4%>%slice(1:40)
p%>%ggplot(aes(reorder(title,n),n,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("Average ratings 3<=r<4 and n >10000")+xlab('movie')+ylab('Number of ratings')+theme_classic()
```
#I observed the movies that had a mean rating between 2 and 3 and the number of movies that had a rating that was greater than 5000
```{r, message=FALSE}
avg_between2_3<-edx%>%group_by(title)%>%summarize(mean_rating=mean(rating,n=n())%>%filter(n>5000,(mean_rating>=2 & mean_rating<3))%>%arrange(desc(n,mean_rating))
avg_between2_3%>%ggplot(aes(reorder(title,n),n,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("Average ratings 2<=r<3 and n>5000")+xlab('Movie')+ylab('Number of Ratings')+theme_classic()
```
#Movies that have fewer than 1000 ratings, movie ratings that are greater than 500, and movie ratings that have a rating of less than 2
```{r,message=FALSE}
avg_rating_less_than_2<-edx%>%group_by(title)%>%summarize(mean_rating=mean(rating),n,n=n())%>%filter(n>500,mean_rating<2)%>%arrange(desc(n,mean_rating))
avg_rating_less_than_2%>%ggplot(aes(reorder(title, n),n,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("Average rating<2")+xlab('Movie')+ylab('Number of Ratings')+theme_classic()
```
#Make the calculation for the least squares for the movieid. Here, we are observing which movies have a greater than average rating and which movies have a larger number of ratings for them.
```{r,message=FALSE}
p<-mean(edx$rating)
edx%>%group_by(title)%>%summarize(b_i=mean(rating-p),n,n=n())%>%filter(b_i>0.5,n>10000)%>%ggplot(aes(reorder(title,b_i),b_i,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("")+xlab("Movie Title")+ggtitle("Movie Rating-p\n for Number of Ratings>10000")+theme_classic()
```
#The regular averages for the movies in the data
```{r,message=FALSE}
movie_avgs<-edx%>%group_by(movieid)%>%summarize(b_i=mean(rating-p))
movie_reg_avgs<-edx%>%group_by(movieid)%>%summarize(b_i=sum(rating-p)/(n()+1),n_i=n())
movie_titles<-edx%>%select(movieid,title)%>%distinct()
edx_with_avgs<-edx%>%group_by(title,movieid)%>%summarize(n=n())%>%left_join(movie_reg_avgs,by="movieid")%>%arrange(desc(b_i,n))
edx_with_avgs%>%filter(n>15000)%>%ggplot(aes(reorder(title,b_i),b_i,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("")+xlab("movie title")+ggtitle("Regularized Averages\nnumber of ratings>20000")+theme_classic()
```
#Movies that have a rating of less than 2 and their regular averages
```{r,message=FALSE}
head(edx_with_avgs)
b<-edx_with_avgs%>%arrange(b_i)%>%filter(b_i<-2)%>%arrange((b_i))
b
b%>%ggplot(aes(reorder(title,b_i),b_i,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("")+xlab("movie title")+ggtitle("Regularized movie averages b_i<-2")+theme_classic()
```
#Movies that contain a regular average of less than 0 and have ratings greater than 1000
```{r,message=FALSE}
edx_with_avgs%>%filter(n>10000,b_i<0.0)%>%ggplot(aes(reorder(title,b_i),b_i,fill=n))+geom_bar(stat="identity")+coord_flip()+scale_fill_distiller(palette="PuBuGn")+ggtitle("")+xlab("movie title")+ggtitle("Regularized movie averages\n number of ratings>20000")+theme_classic()
```
#Observe the correlation pattern between the age of the movie, movieid, userid, ratings, and the quantity of ratings. 
```{r}
n_movies_ratings<-edx_with_title_dates%>%group_by(movieid)%>%summarize(n=n())
avg_movie_rating<-edx_with_title_dates%>%group_by(movieid)%>%summarize(avg_movie_rating=mean(rating))
corr_dat<-edx_with_title_dates%>%select(rating,userid,movieid,age_of_movie,year_rated,premier_date,rating_date_range)%>%leftjoin(n_movies_ratings,by="movieid")%>%left_join(avg_movie_rating,by="movieid")
head(corr_dat)
```
#Observe and graphically display the correlation
```{r,message=FALSE}
temp<-corr_dat%>%select(one_of("movieid","userid","rating","age_of_movie","year_rated","premier_date","n","rating_date_range","avg_movie_rating"))%>%as.matrix()
A<-corr(temp,use="pairwise.complete.obs")
corrplot(A,order="hclust",addrect=2,type="lower",col=brewer.pal(n=8,name="RdBu"))
```
#Does the age of the movie have any effect?
```{r,message=FALSE}
corr_by_age_of_movie<-corr_dat%>%filter((age_of_movie>20)&(age_of_movie<70))
temp<-corr_by_age_of_movie%>%select(one_of("rating","userid","movieid","n","premier_date","rating_date_range","age_of_movie","avg_movie_rating"))%>%as.matrix()
A<-cor(temp,use="pairwise.complete.obs")
corrplot(A,order="hclust",addrect=2,type="lower",col=brewer.pal(n=8,name="RdBu"))
```
#Is there any correlation between the average movie ratings and the quantity of ratings
```{r,message=FALSE}
get_cor<-function(df){
  a<-cor(df$x,df$y,use="pairwise.complete.obs");bx<-substitute(italic(r)==cor,list(cor=format(a,digits=2)))
  as.character(as.expression(eq));
}
corr_dat%>%ggplot(aes(n,avg_movie_rating))+stat_bin_hex(bins=50)+scale_fill_distiller(palette="Spectral")+stat_smooth(method="lm",color="orchid",size=7)+ylab("Average movie rating")+xlab("Number of ratings")
```
#Is there any effect with movie ratings and age?
```{r,message=FALSE}
corr_dat%>%ggplot(aes(age_of_movie,avg_movie_rating))+stat_bin_hex(bins=50)+scale_fill_distiller(palette="Spectral")+stat_smooth(method="lm",color="orchid",size=1)+annotate("text",x=75,y=0.9,label=get_cor(data.frame(x=corr_by_age_of_movie$age_of_movie,y=corr_by_age_of_movie$avg_movie_rating)),parse=TRUE,color="orchid",size=7)+ylab("Average movie rating")+xlab("Age of movie")
```
#RMSE 
```{r,message=FALSE}
RMSE<-function(true_ratings,predicted_ratings){
  sqrt(mean(true_ratings-predicted_ratings)^2))}

}
lambdas<-seq(0,5,0.5)
rmses<-sapply(lambdas,function(1){
  r<-mean(edx_with_title_dates$rating)
  bi<-edx_with_title_dates%>%group_by(movieid)%>%summarize(bi=sum(rating-r)/(n=n()+1))
  ba<-edx_with_title_dates%>%left_join(bi,by="movieid")%>%group_by(userid)%>%summarize(ba=sum(rating-bi-r)/(n()+1))
  predicted_ratings<-edx_with_title_dates%>%left_join(bi,by="movieid")%>%left_join(ba,by="userid")%>%mutate(pred=r+bi+ba)%>%.$pred
  return(RMSE(predicted_ratings,edx_with_title_dates$rating))})
qqplot(lambdas,RMSEs)
lambdas(which.min(RMSEs))
```

}